---
title: "Bombus pesticide exposure: gene expression in head"
output: prepare_data_for_euler_plot.html
author: Joe Colgan, Yannick Wurm http://wurmlab.com
---

## Introduction
This script takes tab-delimited text files as input, which contain the output results of differential expression analyses using the R package, DESeq2. Differential expression analysis was performed by DESeq2 for queens and workers exposed to the neonicotinoids, clothianidin and imidacloprid, using two input data sets for each caste. One data set consisted of estimated counts generated by the pseudoaligner, Kallisto, while the second data set consisted of actual counts generated from genome-level alignments performed by HISAT2 and gene-level read counts extracted using HTSeq. 

The output of this script is the generation of Venn diagram plots.

The first step is to load libraries required for the present analysis. 

```{r, message=FALSE}
# Load libraries; install from scratch if needed
libraries <- c("ggplot2", "lintr", "ggpubr", "hash", "eulerr")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        biocLite(pkgs = lib)
        library(lib, character.only = TRUE )
    }
}
```

## Step One: Load input files:
Read in input files and store as dataframes. 

```{r, message=FALSE}
## Read in input files for queens:
queen_kallisto <- read.table(file = "./input/queen/kallisto_queen_de_genes_for_topGO.txt", 
                             header = TRUE) 
queen_hisat2 <- read.table(file = "./input/queen/hisat2_queen_de_genes_for_topGO.txt", 
                           header = TRUE)

## Read in input files for workers:
worker_kallisto <- read.table(file = "./input/worker/kallisto_worker_de_genes_for_topGO.txt", 
                             header = TRUE) 
worker_hisat2 <- read.table(file = "./input/worker/hisat2_worker_de_genes_for_topGO.txt", 
                           header = TRUE)
```

## Step Two: Subset significantly differentially expressed genes.

```{r, message=FALSE}
## Create list:
sample_list <-   hash(queen_kallisto = queen_kallisto,
                       queen_hisat2 = queen_hisat2,
                       worker_kallisto = worker_kallisto,
                       worker_hisat2 = worker_hisat2)

## Create empty hash to contain subsetted signficant genes:
deseq2_wald_tables <- hash()
## Create a list to populate with the names of significant genes:
de_genes_wald_names <- list()

## For each dataframe, subset the name of signficant genes:
for(name in names(sample_list)){
        print(name)
        ## Subset based on significance:
        deseq2_wald_tables[[name]] <- subset(sample_list[[name]], padj < 0.05)
        print(nrow(deseq2_wald_tables[[name]]))
        ## Subset names of significant genes:
        de_genes_wald_names[[name]] <- row.names(deseq2_wald_tables[[name]])
}
```
## Step Three: Combine input gene lists into a single list:

```{r, message = FALSE}
## Combine lists:
venneu_queen <- euler(combinations = list("Kallisto \nQueen" = de_genes_wald_names$queen_kallisto,
                                    "HISAT2-HTSeq \nQueen" = de_genes_wald_names$queen_hisat2))

venneu_worker <- euler(combinations = list("Kallisto \nWorker" = de_genes_wald_names$worker_kallisto,
                                    "HISAT2-HTSeq \nWorker" = de_genes_wald_names$worker_hisat2))

```

## Step Four: Generate plot:

```{r, message = FALSE}
## Generate plot:
queen_plots <- plot(venneu_queen,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 10))

worker_plots<- plot(venneu_worker,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 10))

## Generate a combined plot:
ggarrange(queen_plots, worker_plots,
          labels = c("A) Queen DE genes", 
                     "B) Worker DE genes"),
          ncol = 1, nrow = 2)

```
## Step Six: Overlap in enriched gene ontology terms
As an additional measure to explore the overlap the results of the differential expression analyses using Kallisto estimated counts and HISAT2-HTSeq actual counts, we explore the overlap in gene ontology terms identified as enriched within each dataset:

```{r, message = FALSE}
## Read in significant gene ontology term enriched for each caste within each dataset:
## Read in queen data:
queen_kallisto_go <- read.table(file = "./input/queen/gene_ontology/kallisto_queen_combined_sig_terms.txt",
                                header = TRUE)

queen_hisat2_go <- read.table(file = "./input/queen/gene_ontology/hisat2_queen_combined_sig_terms.txt",
                                header = TRUE)

## Read in worker data:
worker_kallisto_go <- read.table(file = "./input/worker/gene_ontology/kallisto_worker_combined_sig_terms.txt",
                                header = TRUE)

worker_hisat2_go <- read.table(file = "./input/worker/gene_ontology/hisat2_worker_combined_sig_terms.txt",
                                header = TRUE)

## Create list:
go_list <-   hash(queen_kallisto_go = queen_kallisto_go,
                  queen_hisat2_go = queen_hisat2_go,
                  worker_kallisto_go = worker_kallisto_go,
                  worker_hisat2_go = worker_hisat2_go)

## Create empty hash to contain subsetted signficant terms:
go_tables <- hash()

## Create a list to populate with the names of significant terms:
go_names <- list()

## Ensure that adjusted p-values to subset are numeric:
## For each dataframe, subset the name of signficant genes:
for(name in names(go_list)){
        print(name)
        ## Subset based on significance:
        go_list[[name]]$weight_ks <- as.character(unlist(go_list[[name]]$weight_ks))
        go_tables[[name]] <- subset(go_list[[name]], weight_ks < 0.05)
        print(nrow(go_tables[[name]]))
        ## Subset names of significant genes:
        go_names[[name]] <- go_tables[[name]]$GO.ID
}
```
## Step Seven: Combine input term lists into a single list:

```{r, message = FALSE}
## Combine lists:
venneu_all_go <- euler(combinations = list("Kallisto \nQueen" = go_names$queen_kallisto_go,
                                    "HISAT2-HTSeq \nQueen" = go_names$queen_hisat2_go,
                                    "Kallisto \nWorker" = go_names$worker_kallisto_go,
                                    "HISAT2-HTSeq \nWorker" = go_names$worker_hisat2_go))

venneu_queen_go <- euler(combinations = list("Kallisto \nQueen" = go_names$queen_kallisto_go,
                                    "HISAT2-HTSeq \nQueen" = go_names$queen_hisat2_go))


venneu_worker_go <- euler(combinations = list("Kallisto \nWorker" = go_names$worker_kallisto,
                                    "HISAT2-HTSeq \nWorker" = go_names$worker_hisat2))

```

## Step Eight: Generate plots:

```{r, message = FALSE}
## Generate plot:
all_go_plots <- plot(venneu_all_go,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 10))

## Generate plot for queens:
queen_go_plots <- plot(venneu_queen_go,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 10))

## Generate plot for workers:
worker_go_plots<- plot(venneu_worker_go,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 10))

## Generate a combined plot:
ggarrange(queen_go_plots, worker_go_plots,
          labels = c("A) Queen GO terms", 
                     "B) Worker GO terms"),
          ncol = 1, nrow = 2)

```
