---
title: "Bombus pesticide exposure: gene expression in head"
output: GO_enrichment_3treatments.html
---
# Introduction 
This script is for Gene Ontology (GO) analysis of differentiallly expressed genes to explore the biological processes and molecular functions associated with DE genes, using the R package 'TopGO' (https://www.bioconductor.org/packages/3.7/bioc/vignettes/topGO/inst/doc/topGO.pdf).
Input files should be a .txt file of 'genes of interest' generated from differential gene expression analysis, using the R package DESeq2 and a database of GO terms mapped to the genome of interest. 
This script will prepare data for GO analysis and create a 'TopGO object' from which enrichment tests can be performed to explore GO terms significantly enriched within the dataset. 
This script produces a results table of significantly enriched GO terms.

```{r}
# Load libraries; install from scratch if needed
libraries <- c("topGO")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        biocLite(pkgs = lib)
        library(lib, character.only = TRUE )
    }
}
```

## Step One: Input files and define objects for topGO

We use a list of significantly differentially expressed genes and an input database of GO terms.
```{r}
# Set input files:
genes_of_interest <- read.table("de_genes_for_topGO.txt", header=T)

## Sort by adjusted p-value:
genes_of_interest_sorted <- genes_of_interest[order(genes_of_interest$adjusted_pvalue), ]
## Remove na values:
genes_of_interest_sorted_filtered <- genes_of_interest_sorted[!is.na(genes_of_interest_sorted$adjusted_pvalue),]

# Read in GO annotations
geneID2GO <- readMappings(file = "blast2go_mapping_20171223_1007.input_database_for_topGO.txt")

# Define gene 'universe'
gene_universe <- names(geneID2GO) 

## Create genelist
geneList <-genes_of_interest_sorted_filtered$adjusted_pvalue
names(geneList) <- genes_of_interest_sorted_filtered$locus

# Define a cut-off:
cutoff_for_top_fivepercent <- quantile(x = geneList, probs = 0.05)

# Tell TOPGO where these interesting genes appear in gene universe vector
#gene_list <- factor(as.integer(gene_universe %in% genes_of_interest))
#names(gene_list) <- gene_universe

```

## Step Two: Create topGO data object, using BP for 'Biological Processes'
We create a topGO object using GO terms associated with 'Biological Processes'

```{r}
## Variables to define - 'BP' or 'MF'
GO_term_type <- 'BP'

# build the GOdata object in topGO
myGOdata <- new("topGOdata", 
                description="My project", 
                ontology=GO_term_type, 
                #geneSel = function(x) {return(x=TRUE)}, # Run just for ks test
                geneSel = function(x) {return(x < cutoff_for_top_fivepercent)}, # only used for fisher test
                allGenes=gene_list,  
                gene2GO = geneID2GO, 
                annot = annFUN.gene2GO, 
                nodeSize=100)
```

## Step Three: Run statistical tests for GO term enrichment
We perform three statistical tests:
1. A ks test using the topGO 'elim' algorithm
2. A ks test using the topGO 'weight01' algorithm
3. A Fisher test using the topGO 'weight01' algoritm

We combine the output of each test. 
We filter out enriched terms.

```{r}
## Calculate ks test using 'elim' algorithm
resultElimKs <- runTest(myGOdata, algorithm="elim", statistic="ks")

## Calculate ks test using 'weight01' algorithm:
resultWeightKs <- runTest(myGOdata, algorithm="weight01", statistic="ks")

## Calculate fisher exact test using 'weight01' algorithm:
resultWeightFisher <- runTest(myGOdata, algorithm="weight01", statistic="fisher")

## Combine results from statistical tests:
allRes <- GenTable(myGOdata, 
                   kselim = resultElimKs, 
                   weight = resultWeightKs, 
                   weightfisher =resultWeightFisher,
                   orderBy = "weight", 
                   topNodes = 2038) 

# Filter out calls with significance higher than expected:
allRes <- allRes[which(allRes$Significant > allRes$Expected), ]

## Print to console one of the GO terms of interest:
print(showGroupDensity(myGOdata, 
                       "GO:0044707", 
                       ranks = TRUE, 
                       rm.one=FALSE))

```













