--- 
title: "Bombus pesticide exposure: gene expression in head"
output: plot_DEGs_heatmap.html
author: Isabel Fletcher, Joe Colgan, Yannick Wurm http://wurmlab.com
--- 

## Introduction:

This script is for exploratory analysis of transcript quantification data and results of differential gene expression analysis.
Input files should be R objects saved and generated from previous scripts of DE analysis, including the object resulting from using tximport() on kallisto files, the significant transcript abundances selected from this object and a results table of significant genes, generated using DESeq2. 
This script will organise input data for generating heatmaps of significantly differentially expressed genes, based on statistical significance and log fold changes in expression. This script produces a heatmap that can be exported. 

```{r, message=FALSE}
# Load libraries; install from scratch if needed
libraries <- c("ggplot2", "gplots", "heatmap3",
               "ggfortify", "rentrez",
               "reshape2", "scales", "ggpubr",
               "grid", "XML")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        library(lib, character.only = TRUE )
    }
}

## Load custom function:
source("description_from_LOCid.R")
source("helper_functions_for_heatmaps.R")

## Assign path for output:
output <- "results/heatmap_figure"
dir.create(output, recursive = TRUE)
```

## Step One: Load files containing raw count estimates and differentially expressed genes:
At present this script only plots heatmaps for one pesticide and one control treatment. 
The script generates heatmap plots for two castes: workers and queens.

```{r, message = FALSE}
## Define treatments for plotting:
pesticide <- "clothianidin"
control   <- "control"

## Define castes for plotting:
castes <- c("workers", "queens")

## Generate a function for plotting by caste:
## Combine plot:
ggarrange(caste_plotter(castes[1]), caste_plotter(castes[2]),
                           align = "hv",
                           nrow = 2, ncol = 1,
                           heights = c(3.0, 1.0),
                           common.legend = TRUE,
                           legend = "top")

## Save output:
ggsave("results/heatmap_figure/combined_worker_queen_heatmap.pdf",
       width = 20, height = 18)
```
