---
title: "Bombus pesticide exposure: gene expression in head"
output: plot_go_barchart.html
author: Isabel Fletcher, Joe Colgan, Yannick Wurm http://wurmlab.com 
---

## Introduction:  

This script is for visualisation of the output of gene ontology (GO) analyses using the R package TopGo.
This script creates a barchart of the number of significant genes annotated to a particular GO term, with an associated p-value. This script takes an input of .csv files containing an output table of GO analyses.

```{r, message = FALSE}
# Load libraries; install from scratch if needed
libraries <- c("ggplot2", "lattice", "ggpubr")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        library(lib, character.only = TRUE )
    }
}

## Create output directory:
dir.create("results")
```

## Step One: Load input files
At present, the script only analyses go terms enriched within clothianidin-exposed bumblebees (both castes). 

```{r, message = FALSE}
## Load helper functions:
source("./barchart_helper_functions.R")

## Define pesticide and control treatments:
pesticide <- "clothianidin"
pesticide_abbrev <- substr(pesticide, start = 1, stop = 3)
control   <- "control"

## Define the two bumblebee castes:
castes <- c("worker", "queen")

## Plot:
ggarrange(caste_barchart_plotter(castes[1]),
          caste_barchart_plotter(castes[2]),
          nrow = 2, ncol = 1,
          labels = c("A", "B"),
          align = "hv",
          heights = c(1.5, 1.8),
          font.label = list(size = 10))

## Save picture:
ggsave(paste("results/", pesticide, "_go_terms_barchart.pdf", sep = ""),
       height = 20, width = 18)
```
