---
title: "Bombus pesticide exposure: gene expression in head"
output: run_kallisto_hisat2_go_term_comparison.html
author: Joe Colgan, Yannick Wurm http://wurmlab.com
---

## Introduction
This script takes tab-delimited text files as input, which contain significant gene ontology terms enriched genes as identified by DESeq2.  
The present script takes input for both bumblebee castes (queens and workers) and generates a euler graph for each.
These euler graphs are saved as output as a .png file.  

The first step is to load libraries required for the present analysis. 

```{r, message=FALSE}
# Load libraries; install from scratch if needed
libraries <- c("ggplot2", "lintr", "ggpubr", "hash", "eulerr")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        biocLite(pkgs = lib)
        library(lib, character.only = TRUE )
    }
}
```

## Step One: Load input files:
Read in input files and store as dataframes. 

```{r, message = FALSE}
## Read in significant gene ontology term enriched for each caste within each dataset:
## Read in input files for queens:
queen_kallisto <- read.table(file = "./input_for_euler/queen/kallisto_combined_terms.txt",
                             header = FALSE,
                             col.names = "ids")$ids
queen_hisat2 <- read.table(file = "./input_for_euler/queen/hisat2_combined_terms.txt",
                             header = FALSE,
                             col.names = "ids")$ids

## Read in input files for workers:
worker_kallisto <- read.table(file = "./input_for_euler/worker/kallisto_combined_terms.txt",
                             header = FALSE,
                             col.names = "ids")$ids
worker_hisat2 <- read.table(file = "./input_for_euler/worker/hisat2_combined_terms.txt",
                           header = FALSE,
                             col.names = "ids")$ids
```

## Step Two: Create combinations for plotting:

```{r, message = FALSE}
## Combine lists:
venneu_queen <- euler(combinations = list("Kallisto \nQueen" = queen_kallisto,
                                    "HISAT2-HTSeq \nQueen" = queen_hisat2))

venneu_worker <- euler(combinations = list("Kallisto \nWorker" = worker_kallisto,
                                    "HISAT2-HTSeq \nWorker" = worker_hisat2))

```

## Step Three: Generate plot:

```{r, message = FALSE}
## Generate plot:
queen_plots <- plot(venneu_queen,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 12))

worker_plots <- plot(venneu_worker,
  quantities = TRUE,
  edges = TRUE,
  labels = list(fontsize = 12))

## Generate a combined plot:
ggarrange(queen_plots, worker_plots,
          labels = c("A) Queen GO terms",
                     "B) Worker GO terms"),
          ncol = 1, nrow = 2)

## Save picture:
ggsave("caste_gene_ontology_euler_plot.png",
       height = 10, width = 8)
```